---
title: "BIOINF3005/7160: Transcriptomics Applications"
subtitle: "Course Homepage"
output: 
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = FALSE,
    message = FALSE,
    warning = FALSE
    )
```

```{r}
library(tidyverse)
library(lubridate)
library(pander)
panderOptions("table.style", "rmarkdown")
panderOptions("table.split.table", Inf)
panderOptions("missing", "")
```

```{r}
pracs <- tribble(
    ~Week, ~`Week Starting`, ~Lecture, ~Lecturer, ~LecUrl, ~Practical, ~PracUrl,
    1, "2020-03-02", "Introduction", "Stephen Pederson", "", "Introduction to R and `rmarkdown`", "practicals/Practical1_IntroR.html",
    2, "2020-03-09", "**Adelaide Cup Day**", "No Lecture", "", "Importing Data with the `tidyverse`", "",
    3, "2020-03-16", "Quantification Strategies", "Stephen Pederson", "", "Data Visualisation with `ggplot2`", "",
    4, "2020-03-23", "Microarray Technology", "Stephen Pederson", "", "Using Git with RStudio", "",
    5, "2020-03-30", "Statistical Techniques", "Stephen Pederson", "", "Basic Statistics in R", "",
    6, "2020-04-06", "RNA-Seq", "Zhipeng Qu", "", "Alignments and Visualisation", "",
    NA_integer_, "2020-04-13", "**Mid-Semester Break**", "No Lecture", "", "", "",
    NA_integer_, "2020-04-20", "**Mid-Semester Break**", "No Lecture", "", "", "",
    7, "2020-04-27", "Alignment and Quantification", "Nathan Watson-Haigh", "", "Introduction to `limma` and `edgeR`", "",
    8, "2020-05-04", "Statistics for RNA-Seq", "Stephen Pederson", "", "Introduction to `limma` and `edgeR`", "",
    9, "2020-05-11", "Enrichment Analysis", "Stephen Pederson", "", "Enrichment Strategies", "",
    10, "2020-05-18", "Single Cell RNA-Seq", "Stephen Pederson", "", "WGCNA", "",
    11, "2020-05-25", "Network Analysis", "Nathan Watson-Haigh", "", "WGCNA", "",
    12, "2020-06-01", "Transcriptome Assembly", "Terry Bertozzi", "", "", "",
    13, "2020-06-07", "Challenges in RNA-Seq", "Stephen Pederson", "", "", ""
)
```



# Timetable

- **Lectures**
    - Monday 9:10am - 10:00am, Ingkarni Wardli, Room B18
- **Practicals**
    - Wednesday 9:10am - 11:00am, Johnson Building, Room 111
    - Friday 9:10am - 11:00am, Johnson Building, Room 111

Lectures for this course are held on a Monday, and will clash with **two Public Holidays** in Semester 1.

1. Week 2, March 9^th^, Adelaide Cup Day: **No Lecture in Week 2**
2. Week 13, June 7^th^, Queen's Birthday: An alternative lecture time may be found for this week. This will be announced closer to the date.


```{r}
pracs %>%
    mutate(
        `Week Starting` = as_date(`Week Starting`), 
        Lecture = case_when(
            !grepl("\\*\\*", Lecture) ~ paste0("[", Lecture, "](", LecUrl, ")", "<br>(", Lecturer, ")"),
            grepl("\\*\\*", Lecture) ~ paste0(Lecture, " (", Lecturer, ")")
        ),
        Practical = paste0("[", Practical, "](", PracUrl, ")")
    ) %>%
    dplyr::select(starts_with("Week"), Lecture, Practical) %>%
    pander(
        justify = "rrll"
    )
```



# Assessments

```{r}
ass <- tribble(
    ~Url, ~Topic, ~`Due Date`,
    "", "", ""
) %>%
    mutate(
        Assessment = paste("Assessment", seq_len(nrow(.)) - 1)
    )
```


## Assessment Tasks

```{r}
ass %>%
    mutate(
        Assessment = paste0("[", Assessment, "](", Url, ")")
    ) %>%
    dplyr::select(Assessment, Topic, `Due Date`) %>%
    pander(
        justify = "llr"
    )
```


## Assessment Checklist

Have you:

- [ ] Answered all the questions?
- [ ] Followed naming conventions for Assessments?
- [ ] Checked that you have not breached the [Academic Honesty Policy](http://www.adelaide.edu.au/policies/230/).
- [ ] Identified the work as yours?
	- Emails should have the course and assessment task names.
	- Documents should be named with your name, the course name and the assessment task.
	- Printed documents should have you name and the course and assessment task in the text/footer/header.
- [ ] Used appropriate electronic communication with assessors?
	- Emails should have a meaningful subject.
- [ ] Handed in the assignment before the due time (see MyUni)?

# Useful Links

[How To Ask Questions The Smart Way](http://www.catb.org/esr/faqs/smart-questions.html)

[How to write a good bug report](https://musescore.org/en/developers-handbook/how-write-good-bug-report-step-step-instructions)
