---
title: "Transcriptomics Applications"
subtitle: "Assignment 1"
output: 
  html_document: 
    toc: yes
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = FALSE, 
    warning = FALSE,
    message = FALSE
)
```

```{r}
library(tidyverse)
```


# Instructions

## Submission Format [3 marks]

This is assignment is due by 5pm, **Tuesday 24th March**.

- Submissions must be made as a *zip archive* containing 3 files:
    1. Your **source R Markdown Document** (with an Rmd suffix)
    2. A **compiled pdf**, showing all code
    3. The **signed cover sheet** as required by the University
- All 3 file names **within the zip archive** must start with your student number.
However the name of the zip archive is not important as myUni will likely modify this during submission. [See here for help creating a zip archive](#creating-a-zip-archive)
    
*All questions are to be answered on the same R Markdown / PDF, regardless of if they require a plain text answer, or require execution of code.*

We **strongly advise** working in the folder `~/transcriptomics/assignment1` on your virtual machine.
Using an R Project for each individual assignment is also strongly advised.

## Creating a zip archive {.tabset}

### On Your VM

If all files required for submission *are contained on your VM*:

1. Select all three files using the **Files** pane in R Studio
2. Click export
3. They will automatically be placed into a single zip archive. Please name this in whatever informative name you decide is suitable, but it should contain the suffix `.zip`

### Windows

If all files are on your on your *local Windows machine*:

1. Using File Explorer, enter the folder containing all 3 files
2. Select all files simultaneously by using <kbd>Ctrl</kbd> + Click
3. Right click on one of the files and select `Send to > Compressed (zipped) folder`
4. Rename as appropriate, ensuring the archive ends with the suffix `.zip`

### Mac OS

If all the files are on your *local macOS machine`:

1. Locate the items to zip in the Mac Finder (file system)
2. Right-click on a file, folder, or files you want to zip
3. Select “Compress Items”
4. Find the newly created .zip archive in the same directory and name as appropriate

# Questions

```{r}
marks <- tibble(q = 0, marks = 3)
q <- 1
```


### Question `r q` [8 marks]

```{r}
marks <- marks %>%
    bind_rows(
        tibble(q = q, marks = 8)
    )
```


Choose two different RNA types and contrast them with each other.
Aspects to consider may be method of transcription, cellular location, post-transcriptional processing, biological function or any other aspect which you determine to be important

```{r}
q <- q + 1
```

### Question `r q` [3 marks]

```{r}
marks <- marks %>%
    bind_rows(
        tibble(q = q, marks = 3)
    )
```


In R, you will commonly encounter 3 types of 'unexpected' output.
1) Errors, 2) Warnings and 3) Messages.
Describe the role of each of these and how to interpret them.

```{r}
q <- q + 1
```

### Question `r q` [6 marks]

```{r}
marks <- marks %>%
    bind_rows(
        tibble(q = q, marks = 6)
    )
```


Two possible definitions of a gene are given by the high-profile journal [Nature](https://www.nature.com/scitable/definition/gene-29/) and the US [National Institute of Health](https://ghr.nlm.nih.gov/primer/basics/gene).
Discuss the limitations of these definitions, giving particular consideration to promoters and protein products which arise from multiple distinct locations within the genome.
Two interesting discussion on this subject are available in [this paper](https://www.genetics.org/content/205/4/1353) and [this lecture](https://youtu.be/TiaQG0UZPAU).
Feel free to use these resources, or find your own.

```{r}
q <- q + 1
```

### Question `r q` [ marks]

<!-- Provide a unique set of filenames for each student and ask them to form a tibble extracting the key information. -->
<!-- Also ask them to summarise the metadata, by merging with additional information -->

# Total `r sum(marks$marks)` marks

```{r}
write_csv(marks, here::here("assignments", "A1.csv"))
```

