---
title: "Transcriptomics Applications"
subtitle: "Assignment 2"
output: 
  html_document: 
    toc: yes
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = FALSE, 
    warning = FALSE,
    message = FALSE
)
```

```{r}
library(tidyverse)
```

# Instructions

## Submission Format [3 marks]

This is assignment is due by 5pm, **Friday 8^th^ May**.

- Submissions must be made as a *zip archive* containing 2 files:
    1. Your **source R Markdown Document** (with an Rmd suffix)
    2. A **compiled pdf**, showing all code
- All file names **within the zip archive** must start with your student number.
However the name of the zip archive is not important as myUni will likely modify this during submission. [See here for help creating a zip archive](#creating-a-zip-archive)
    
*All questions are to be answered on the same R Markdown / PDF, regardless of if they require a plain text answer, or require execution of code.*

**Marks directly correspond to the amount of time and effort we expect for each question, so please answer with this is mind.**

We **strongly advise** working in the folder `~/transcriptomics/assignment2` on your virtual machine.
Using an R Project for each individual assignment is also strongly advised.

## Creating a zip archive {.tabset}

### On Your VM

If all files required for submission *are contained on your VM*:

1. Select both files using the **Files** pane in R Studio
2. Click export
3. They will automatically be placed into a single zip archive. Please name this in whatever informative name you decide is suitable, but it should contain the suffix `.zip`

### Windows

If all files are on your on your *local Windows machine*:

1. Using File Explorer, enter the folder containing both files
2. Select all files simultaneously by using <kbd>Ctrl</kbd> + Click
3. Right click on one of the files and select `Send to > Compressed (zipped) folder`
4. Rename as appropriate, ensuring the archive ends with the suffix `.zip`

### Mac OS

If all the files are on your *local macOS machine`:

1. Locate the items to zip in the Mac Finder (file system)
2. Right-click on a file, folder, or files you want to zip
3. Select “Compress Items”
4. Find the newly created .zip archive in the same directory and name as appropriate


# Questions

```{r}
marks <- tibble(q = 0, marks = 3)
q <- 1
```

## Question `r q` [8 marks]

```{r}
marks <- marks %>%
    bind_rows(
        tibble(q = q, marks = 8)
    )
```

Two common strategies for RNA Seq library preparation are the depletion of rRNA molecules or the preferential amplification of poly-adenylated RNA.
Briefly contrast these two approaches, describing their respective strengths and limitations.

```{r}
q <- q + 1
```

## Question `r q` [10 marks]

```{r}
marks <- marks %>%
    bind_rows(
        tibble(q = q, marks = 10)
    )
```

A common alignment and quantification workflow is to align an RNA Seq sample to a reference genome and count reads which align to each gene, as defined in a `gtf` file.

a. Describe two important considerations during the quantification step, and how these may differ for polyA and total RNA libraries?
b. What advantages and disadvantages does this approach offer in comparison to using pseudo-alignment to a reference transcriptome


```{r}
q <- q + 1
```









# Total: `r sum(marks$marks)` marks

```{r}
write_csv(marks, here::here("assignments", "A3.csv"))
```
