---
title: "Transcriptomics Applications"
subtitle: "Assignment 5"
output: 
  html_document: 
    toc: yes
    toc_depth: 2
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = FALSE, 
    warning = FALSE,
    message = FALSE
)
```

```{r}
library(tidyverse)
```

# Instructions

## Submission Format [3 marks]

This is assignment is due by 5pm, **Tuesday 9^th^ June**.

- Submissions must be made as a *zip archive* containing 2 files:
    1. Your **source R Markdown Document** (with an Rmd suffix)
    2. A **compiled pdf**, showing all code
- All file names **within the zip archive** must start with your student number.
However the name of the zip archive is not important as myUni will likely modify this during submission. [See here for help creating a zip archive](#creating-a-zip-archive)
    
*All questions are to be answered on the same R Markdown / PDF, regardless of if they require a plain text answer, or require execution of code.*

**Marks directly correspond to the amount of time and effort we expect for each question, so please answer with this is mind.**

We **strongly advise** working in the folder `~/transcriptomics/assignment4` on your virtual machine.
Using an R Project for each individual assignment is also strongly advised.

## Creating a zip archive {.tabset}

### On Your VM

If all files required for submission *are contained on your VM*:

1. Select both files using the **Files** pane in R Studio
2. Click export
3. They will automatically be placed into a single zip archive. Please name this in whatever informative name you decide is suitable, but it should contain the suffix `.zip`

### Windows

If all files are on your on your *local Windows machine*:

1. Using File Explorer, enter the folder containing both files
2. Select all files simultaneously by using <kbd>Ctrl</kbd> + Click
3. Right click on one of the files and select `Send to > Compressed (zipped) folder`
4. Rename as appropriate, ensuring the archive ends with the suffix `.zip`

### Mac OS

If all the files are on your *local macOS machine`:

1. Locate the items to zip in the Mac Finder (file system)
2. Right-click on a file, folder, or files you want to zip
3. Select “Compress Items”
4. Find the newly created .zip archive in the same directory and name as appropriate


# Questions

```{r}
marks <- tibble(q = 0, marks = 3)
q <- 1
```

## Question `r q` [5 marks]

```{r}
marks <- marks %>%
    bind_rows(
        tibble(q = q, marks = 5)
    )
```

A *Module Eigengene* (ME) is one of the fundamental concepts of gene co-expression network analysis using WGCNA. 

i. Is ME one of the genes in a co-expression module? Explain your reasoning
ii. Describe the roles (usage) of ME in gene co-expression network analysis

```{r}
q <- q + 1
```

## Question `r q` [4 marks]

```{r}
marks <- marks %>%
    bind_rows(
        tibble(q = q, marks = 4)
    )
```

We have discussed that the key point of WGCNA analysis is to discover biological significance.
Describe how can we build the link between our co-expression network analysis and biological significance.

```{r}
q <- q + 1
```


## Question `r q` [4 marks]

```{r}
marks <- marks %>%
    bind_rows(
        tibble(q = q, marks = 4)
    )
```

Bulk RNA-seq and scRNA-seq differ in many ways.
Describe two common aspects shared between the two approaches, and provide details about two key differences between the two.

```{r}
q <- q + 1
```


## Question `r q` [12 marks]

```{r}
marks <- marks %>%
    bind_rows(
        tibble(q = q, marks = 12)
    )
```

Use the given [gene expression dataset](data/WGCNA_rawCounts.tsv) and corresponding [clinical data](data/WGCNA_clinical_data.tsv) to identify co-expression module(s) with the strongest correlation with clinical data. 
The gene expression dataset includes raw counts for human reference genes across 37 samples (skin cancer patients), and the clinical data includes diagnosed cancer stage status of 37 patients. 
The results should minimally include:

i. A figure showing which and how the power beta is selected.
ii. A gene dendrogram and coloured co-expression modules detection using dynamic tree cut

```{r}
q <- q + 1
```

## Question `r q` [12 marks]

```{r}
marks <- marks %>%
    bind_rows(
        tibble(q = q, marks = 12)
    )
```

Using the [feature-barcode matrix] provided, perform the following steps.
Please note that undetectable genes have already been removed from this dataset.

i. Identify and remove low quality cells, explaining why you are performing each step
ii. Perform PCA and generate a plot of the cells using PC1 and PC2
iii. Choose a suitable method and define clusters
iv. Overlay your clusters on either a UMAP or tSNE plot
v. Discuss your clusters and comment on how effective your approach was

```{r}
q <- q + 1
```



# Total: `r sum(marks$marks)` marks

```{r}
write_csv(marks, here::here("assignments", "A5.csv"))
```
