---
title: "BIOINF3005/7160: Transcriptomics Applications"
subtitle: "Reshaping & Plotting Data with the `tidyverse`"
date: "13^th^ March, 2020"
output: 
  html_document:
    toc: yes
    toc_float: yes
---
<script>
function myFunction(i) {
  var x = document.getElementById(i);
  if (x.style.display === "none") {
    x.style.display = "block";
  } else {
    x.style.display = "none";
  }
}
</script>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.align = "center",
    results = "hide"
    )
if (interactive()) setwd(here::here("practicals"))
```


## Introduction

### Outline

In our previous session, we learned how to:

- Import spreadsheet-like data into `R` using `readr`
- Use the `SQL`-like features of `dplyr` to work with our data
- How to print markdown tables from a `tibble`

In this session, we'll look at working with `tibble` objects once again, but this time we'll learn to change the shape of them, and make some amazing looking plots.
After that, we'll try to import some more challenging data.

### Setup

Before we go any further, we need to ensure we're developing our good programming practice.
Make sure you are in your `practical_2` R Project and that all of the directories look correct in your **Files** pane.

The datasets we'll use for today are `topTable.csv` and `cpm.csv`, which will be in `~/data/transcriptomics`.
Please copy these into your `practical_2` directory.

Last time we discovered that setting `message = FALSE` inside the call to `knitr::opts_chunk$set()` helped hide the helpful messages that the `tidyverse` prints for us, so please add that to your setup chunk.
If all goes well, it should look at little like this:

    ```{r setup, include=FALSE}`r ''`
    knitr::opts_chunk$set(
        echo = TRUE,
        message = FALSE
    )
    ```

Now we have our `setup` chunk, let's create a `packages` chunk and load the `tidyverse`.

```{r}
library(tidyverse)
```

Today we'll look at the packages `ggplot2` and `tidyr` which are loaded by default with the `tidyverse` and the second of these is actually where the name came from.


