---
title: "BIOINF3005/7160: Transcriptomics Applications"
subtitle: "Reshaping Data with `tidyr"
date: "13^th^ March, 2020"
output: 
  html_document:
    toc: yes
    toc_float: yes
---
<script>
function myFunction(i) {
  var x = document.getElementById(i);
  if (x.style.display === "none") {
    x.style.display = "block";
  } else {
    x.style.display = "none";
  }
}
</script>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    fig.align = "center",
    results = "hide"
    )
if (interactive()) setwd(here::here("practicals"))
```


## Introduction

### Outline

In our previous session, we learned how to:

- Import spreadsheet-like data into `R` using `readr`
- Use the `SQL`-like features of `dplyr` to work with our data
- How to print markdown tables from a `tibble`

In this session, we'll look at working with `tibble` objects once again, but this time we'll learn to change the shape of them.
After that, we'll try to import some more challenging data.

### Setup

Before we go any further, we need to ensure we're developing our good programming practice.
Make sure you are in your `practical_2` R Project and that all of the directories look correct in your **Files** pane.

You should already have `Enrichment.csv` in your `data` subdirectory, so once you've checked for that, begin a new R Markdown document and save this as `Using_tidyr.Rmd` in your project root directory.
As per usual, delete everything after the initial `setup` chunk that is provided for you by default.

Last time we discovered that setting `message = FALSE` inside the call to `knitr::opts_chunk$set()` helped hide the helpful messages that the `tidyverse` prints for us, so please add that to your setup chunk.
If all goes well, it should look at little like this:

    ```{r setup, include=FALSE}`r ''`
    knitr::opts_chunk$set(
        echo = TRUE,
        message = FALSE
    )
    ```

Now we have our `setup` chunk, let's create a `packages` chunk and load the `tidyverse`.

```{r}
library(tidyverse)
```

Today we'll look at the package `tidyr` which is loaded by default with the `tidyverse` and is actually where the name came from.

The next step is to load the `Enrichment` dataset from the last section, and then we'll be ready to go for today.
A good name for this chunk might be `Enrichment`, or `import`, or `load_Enrichment`.
Whichever makes it clear to you what you're doing.

```{r}
Enrichment <- read_csv("data/Enrichment.csv")
```

```{r, results='markup'}
glimpse(Enrichment)
```

