---
title: "BIOINF3005/7160:<br>Transcriptomics Applications"
subtitle: "Week 3.1: R Object Types"
date: "18^th^ March 2020"
output:
  ioslides_presentation: 
    css: ../custom.css
    fig_caption: yes
    logo: ../lectures/figures/UoA_logo_col_vert.png
    widescreen: yes
---


```{r, loadPackages, echo = FALSE, include = FALSE}
library(knitr)
opts_chunk$set(
  echo = TRUE, include = TRUE, 
  warning = FALSE, message = FALSE, 
  out.width = 800, fig.align = "center"
)
```


# `R` Objects

## Before we start 

1. Create a new R Project `practical_3` (`~/transcriptomics`)
2. Create a new R Markdown:  
    + `File > New File > R_Markdown`
    + Save as `DataTypes.Rmd`

```{r}
library(tidyverse)
```


## Recap Of Weeks 1 & 2

We learned how to:

1. Write reports with `rmarkdown`
2. Import tabular data
3. Look through & summarise a `tibble/data.frame`
4. Use the magrittr (`%>%`)
5. Generate plots with `ggplot2`


## `R` Objects

- Main data type so far has been a `data.frame` 
    - Very much like a spreadsheet
    - `tibble` =  a `data.frame` with pretty wrapping paper
- Each column has **the same type of data**, e.g. `numeric`, `character` etc.
- The columns **can be a different type** to the other columns
- In `R` each column is a `vector`

# Vectors

## Vectors

The key building blocks for `R` objects: **Vectors**

- There is no such thing as a *scalar* in `R`
- Everything is based around the concept of a *vector*

**What is a vector?**

## Vectors | Definition

A vector is one or more values of the **same type**

## Vectors | Examples

A simple vector would be

```{r, echo=FALSE, include=TRUE, results='markup'}
1:10
```

**What type of values are in this vector?**

## Vectors | Examples

Another vector might be

```{r, echo=FALSE, include=TRUE, results='markup'}
c("a", "cat", "video")
```

**What type of values are in this vector?**

## Vectors | Examples

**What type of values are in this vector?**

```{r, echo=FALSE, include=TRUE, results='markup'}
c(742, "Evergreen", "Tce")
```


## The 4 Atomic Vector Types

- *Atomic Vectors* are the building blocks for everything in `R`
- There are four main types
- Plus two we can ignore

(Please start running these examples in your own R Markdown)

## The 4 Atomic Vector Types | Logical Vectors

1. **logical**: Can only hold the values `TRUE` or `FALSE`

```{r}
logi_vec <- c(TRUE, TRUE, FALSE)
logi_vec
```


## The 4 Atomic Vector Types | Integer Vectors

1. logical
2. **integer**: Counts, ranks or indexing positions 

```{r}
int_vec <- 1:5
int_vec
```


## The 4 Atomic Vector Types | Double (i.e. Double Precision) Vectors

1. logical
2. integer
3. **double**: Often (& *lazily*) referred to as `numeric`

```{r, results='hide'}
dbl_vec <- c(0.618, 1.414, 2)
dbl_vec
```

**Why are these called doubles?**

## The 4 Atomic Vector Types | Character Vectors

1. logical
2. integer
3. double
4. **character**

```{r, include=TRUE, results='hide'}
char_vec <- c("blue", "red", "green")
char_vec
```

## The 4 Atomic Vector Types

These are the basic building blocks for all `R` objects

1. logical
2. integer
3. double
4. character

## The 4 Atomic Vector Types

- There are two more rare types we'll ignore:
    - `complex` & `raw`
- All `R` data structures are built on these


## Properties of a vector {.build}

What properties might a vector have?

1. The actual *values*
2. *Length*, accessed by the function `length()`
3. The *type*, accessed by the function `typeof()`
    + Similar but preferable to `class()`
    
## Properties of a vector

What properties might a vector have?

1. The actual *values*
2. *Length*, accessed by the function `length()`
3. The *type*, accessed by the function `typeof()`
4. Any optional & additional *attributes*: `attributes()`
    + Holds data such as `names` etc.

## Working with Vectors

We can *combine* two vectors in `R`, using the function `c()`

```{r}
c(1, 2)
```

- The numbers `1` & `2` were both vectors with `length()` == 1

- We have combined two vectors of length 1, to make a vector of length 2

## Working with Vectors {.build}

**What would happen if we combined two vectors of different types?**

Let's try & see what happens:

```{r, results='hide'}
new_vec <- c(logi_vec, int_vec)
print(new_vec)
typeof(new_vec)
```

## Working with Vectors {.build}

**Q: What happened to the `logical` values?**

Answer: `R` *coerced* them into a common type (i.e. integers).

# Coercion

## Coercion | Discussion Questions {.build}

**What other types could `logical` vectors be coerced into?**

Try using the functions:
`as.integer()`, `as.double()` & `as.character()` on `logi_vec`

## Coercion {.build}

**Can `character` vectors be coerced into `numeric` vectors?**

```{r}
simp_vec <- c("742", "Evergreen", "Terrace")
as.numeric(simp_vec)
```

<div class="red">
```
Warning message:
NAs introduced by coercion 
```
</div>

# Subsetting Vectors

## Subsetting Vectors

One or more elements of a vector can be called using `[]`

```{r, results='hide'}
char_vec
char_vec[2]
char_vec[2:3]
```

## Subsetting Vectors

Double brackets (`[[]]`) can be used to return **single elements** only

```{r, eval =1}
char_vec[[2]]
```


If you tried `char_vec[[2:3]]` you would receive an error message

## Subsetting Vectors

Double brackets (`[[]]`) can be used to return **single elements** only

```{r, eval =1}
char_vec[[2]]
```

If you tried `char_vec[[2:3]]` you would receive an error message

<div class="red">
```
Error in char_vec[[2:3]] : 
  attempt to select more than one element in vectorIndex 
```
</div>

## Subsetting Vectors

If a vector has name attributes, we can call values by name.

Here we'll use the built-in vector `euro`

```{r}
head(euro)
euro["ESP"]
```

## Subsetting Vectors 

*Try repeating the* call-by-name *approach using double brackets*

```{r, results='hide'}
euro["ESP"]
euro[["ESP"]]
```

**What was the difference in the output?**

## Subsetting Vectors 

*Try repeating the* call-by-name *approach using double brackets*

```{r, results='hide'}
euro["ESP"]
euro[["ESP"]]
```

**What was the difference in the output?**

1. Using `[]` returned the vector with the identical structure
2. Using `[[]]` removed the `attributes` & just gave the R Object at that position (i.e. a numeric vector of length 1)

## Subsetting Vectors | Discussion Question

**Is it better to call by position, or by name?**

## Subsetting Vectors | Discussion Question

**Is it better to call by position, or by name?**

Things to consider:

- Which is easier to type on the fly?
- Which is easier to read?
- Which is more robust to undocumented changes in an object?

## Subsetting Vectors | Extracting Multiple Values 

**What is really happening in this line?**

```{r}
euro[1:3]
```

## Subsetting Vectors | Extracting Multiple Values 

**What is really happening in this line?**

```{r}
euro[1:3]
```

We are using the `integer` vector `1:3` to extract values from the `euro` vector

## Subsetting Vectors | Extracting Multiple Values 

```{r}
int_vec
euro[int_vec]
```


## Vector Operations {.build}

We can also combine the above logical test and subsetting

```{r}
dbl_vec
dbl_vec > 1
```
```{r, results='hide'}
dbl_vec[dbl_vec > 1]
```


## Vector Operations {.build}

An additional logical test: `%in%` (read as: "*is in*")

```{r}
dbl_vec
int_vec
```

```{r, results='hide'}
dbl_vec %in% int_vec
```

## Vector Operations 

```{r}
dbl_vec %in% int_vec
```

Returns `TRUE/FALSE` for each value in `dbl_vec` if it **is in** `int_vec`

NB: `int_vec` was coerced silently to a `double` vector

# Matrices

## Matrices

- Vectors are strictly one dimensional and have a `length` attribute.
- A `matrix` is the **two dimensional** equivalent

```{r, results='hide'}
int_mat <- matrix(1:6, ncol = 2)
print(int_mat)
```

## Matrices

- Matrices can only hold **one** type of value
    + i.e. logical, integer, double, character
- Have additional attributes such as `dim()`, `nrow()` `ncol()`
- Can have **optional** `rownames()` & `colnames()`

## Matrices

**Some commands to try:**

```{r, results='hide'}
dim(int_mat)
nrow(int_mat)
typeof(int_mat)
class(int_mat)
attributes(int_mat)
colnames(int_mat)
length(int_mat)
```

**Please ask questions if anything is confusing**

## Matrices

- Use square brackets to extract values by row & column
- The form is `x[row, col]`
- Leaving either `row` or `col` blank selects the entire row/column

```{r, results='hide'}
int_mat[2, 2]
int_mat[1,]
```

**How would we just get the first column?**

## Matrices

**NB: Forgetting the comma will treat the matrix as a single vector running down the columns**

```{r}
int_mat
int_mat[5]
length(int_mat)
```

## Matrices

**NB: Forgetting the comma will treat the matrix as a single vector running down the columns**

```{r}
int_mat[5]
length(int_mat)
```


## Matrices

**Requesting a row or column that doesn't exist is the source of a very common error message**

```{r}
dim(int_mat)
```

```{r, eval=FALSE}
int_mat[5,]
```
<div class="red">
```
Error in int_mat[5, ] : subscript out of bounds
```
</div>

## Matrices

If row/colnames are assigned:

Can also extract values using these instead of by position

## Arrays

Arrays extend matrices to 3 or more dimensions

Beyond the scope of this course, but we just have more commas in the square brackets, e.g.
```{r}
dim(iris3)
```

```{r,eval=FALSE}
dimnames(iris3)
```

## Summary | Homogeneous Data Types

- Vectors, Matrices & Arrays are the basic *homogeneous* data types of `R`
- All are essentially just vectors

# Heterogeneous Data Types

## Heterogeneous Data Types

Summary of main data types in `R`

Dimension   | Homogeneous | Heterogeneous
------------|-------------|--------------
1d          | `vector`    | `list`
2d          | `matrix`    | `data.frame`
3d+         | `array`     |

## Lists

A `list` is a **heterogeneous** vector.

- Each element is an `R` object
- Can be a `vector`, or `matrix`
- Could be another `list`
- Any other `R` object type we haven't seen yet

**These are incredibly common in `R`**

## Lists

Many `R` functions provide output as a list

```{r, results='hide'}
testResults <- t.test(dbl_vec)
class(testResults)
typeof(testResults)
testResults
```

NB: There is a function (`print.htest()`) that tells `R` how to print the results to the Console

## Lists

Explore the various attributes of the object `testResults`

```{r, results='hide'}
attributes(testResults)
```

Compare this with the results from:

```{r, results='hide'}
attributes(euro)
```

## Lists

```{r, results='hide'}
length(testResults)
names(testResults)
```

## Lists

We can call the individual components of a list using the `$` symbol followed by the name

```{r, results='hide'}
testResults$statistic
testResults$conf.int
testResults$method
```

*Note that each component is quite different to the others.*

## Subsetting Lists

A `list` is a `vector` so we can also subset using the `[]` method

```{r, results='hide'}
testResults[1]
typeof(testResults[1])
```

- Using single square brackets returns a `list` 
    - i.e. an object which is a subset of the larger object, but of the *same type*

## Subsetting Lists

Double brackets again retrieve a single element of the vector

- Returns the actual element as the underlying `R` object

```{r, results='hide'}
testResults[[1]]
typeof(testResults[[1]])
```

## Subsetting Lists

We can also use names instead of positions

```{r, results='hide'}
testResults[c("statistic", "p.value")]
testResults[["statistic"]]
```

- `testResults[["statistic"]]` is identical to `testResults$statistic`

## Lists

- Note also the **Environment** Tab in the top right of RStudio
- Click the arrow next to `testResults` to expand the entry
- This is the output of `str(testResults)`

# Data Frames

## Data Frames

Finally!

- These are the *most common type of data* you will work with
- Each column is a `vector`
- Columns can be different types of vectors
- Column vectors **MUST** be the same length


## Data Frames

- Analogous to matrices, but are **specifically for heterogeneous data**
- Have many of the same attributes as matrices
    + `dim()`, `nrow()`, `ncol()`, `rownames()`, `colnames()`
- `colnames()` & `rownames()` are **NOT** optional & are assigned by default

## Data Frames

Let's use `band_members` again

**Try these commands**

```{r, results='hide'}
colnames(band_members)
rownames(band_members)
dim(band_members)
nrow(band_members)
```

## Data Frames

Individual entries can also be extracted using the square brackets

```{r, results='hide'}
band_members[1:2, 1]
```

We can also refer to columns by name (same as matrices)

```{r, results='hide'}
band_members[1:2, "name"]
```


## Data Frames

Thinking of columns being vectors is quite useful

- We can call each column vector of a `data.frame` using the `$` operator

```{r, results='hide'}
band_members$name[1:2]
```

**There is no equivalent for rows!!!**

## Data Frames & Matrices 

- Many `matrix` objects look exactly like a `data.frame`
    - If both rownames and colnames are set
- `tibble` objects are still clearly distinct
    - There is no `tibble` equivalent for matrices
- The easiest way to check is `class(object)`
- If you try `dplyr` functions on a `matrix`, you will get errors
    - Just use `as.data.frame()` to coerce a matrix to a `data.frame`

# Data Frames & Lists

## Data Frames & Lists

**Data frames are actually special cases of lists**

- Each column of a `data.frame` is an element of a `list`
- The element must all be vectors of the **same length**
- Data frames can be treated identically to a `list`
- Have additional subsetting operations and attributes

## Data Frames & Lists

Forgetting the comma, now gives a completely different result to a matrix!

```{r, eval=FALSE}
band_members[1]
```

**Was that what you expected?**

**Try using the double bracket method**

## Data Frames & Lists | More Errors

**What do you think will happen if we type:**

```{r, eval=FALSE}
band_members[5]
```

## Data Frames & Lists | More Errors

**What do you think will happen if we type:**

```{r, eval=FALSE}
band_members[5]
```

<div class="red">
Error: Positive column indexes in `[` must match number of columns:
* `.data` has 2 columns
* Position 1 equals 5
</div>


# Working With `R` Objects

## Vectors | Name Attributes

**How do we assign names?**

```{r}
named_vec <- c(a = 1, b = 2, c = 3)
```

OR we can name an existing vector

```{r}
names(int_vec) <- c("a", "b", "c", "d", "e")
```

## Vectors | Name Attributes

**Can we remove names?**

The `NULL`, or empty, vector in `R` is created using `c()`

```{r, results='hide'}
null_vec <- c()
length(null_vec)
```

## Vectors | Name Attributes

We can use this to remove names

```{r}
names(int_vec) <- c()
```

<!-- *Don't forget to put the names back...* -->

## Lists

Lists can have `names`, but **not** row/colnames

```{r}
my_list <- list(int_vec, dbl_vec)
names(my_list) <- c("integers", "doubles")
```

OR

```{r}
my_list <- list(integers = int_vec, doubles = dbl_vec)
```

## Lists

**What happens if we try this?**

```{r}
my_list$logical <- logi_vec
```

## Data Frames

This is exactly the same as creating lists, but

**The `names` attribute will also be the `colnames()`**

```{r}
my_df <- data.frame(doubles = dbl_vec, logical = logi_vec)
names(my_df) == colnames(my_df)
```

## Data Frames

**What happens if we try to add components that aren't the same length?**

```{r, eval=FALSE}
my_df <- data.frame(
  integers = int_vec, doubles = dbl_vec, logical = logi_vec
  )
```

<div class="red">
Error in data.frame(integers = int_vec, doubles = dbl_vec, logical = logi_vec) : 
  arguments imply differing number of rows: 5, 3
</div>

## Summary

- These are **all of the basic R objects** you will come across
- **Every R object is based on these!**
- All are `S3` objects
- In transcriptomics `S4` objects are very common
    - The same basic structures still underlie `S4` objects

# Logical Tests

## Logical Tests 

- Is Equal To: `==`
- Not equal: `!=`
- And: `&`
- Or: `|`
- Less than: `<`; Greater than: `>`
- Less than or equal to: `<=`;  Greater than or equal to `>=`

## Logical Tests

- `NA` represents a missing value

```{r, results='hide'}
x <- c(1:5, NA)
x == 5
x != 5
x > 3
x > 3 | x == 2
is.na(x)
```

## Logical Tests

With logical vectors, `!` will invert all values

```{r, results='hide'}
logi_vec
!logi_vec
!is.na(x)
!x == 5
```

- Note the precedence of operations in the last one!
- Is there a more transparent way of writing this?

## Logical Tests

A few more challenging tests which may give unexpected results

```{r, results='hide'}
is.integer(x)
x == int_vec
x[!is.na(x)] == int_vec
x[5:1] == int_vec
```

**Did you understand all of these results?**

## Logical Tests

One final and *important* test in `R`

- `%in%` can be read as *is in*

```{r}
"red" %in% char_vec
char_vec %in% "red"
```

<!-- # Factors -->

<!-- ## Factors -->

<!-- A common data type in statistics is a *categorical variable* (i.e. a `factor`) -->

<!-- - Data will be a set of common groups -->

<!-- ```{r} -->
<!-- pet_vec <- c("Dog", "Dog", "Cat", "Dog", "Cat") -->
<!-- ``` -->

<!-- - This is a `character` vector -->

<!-- ## Factors -->

<!-- - We can simply coerce this to a vector of factors -->
<!-- - Categories will automatically be assigned alpha-numerically -->

<!-- ```{r} -->
<!-- pet_factors <- as.factor(pet_vec) -->
<!-- pet_factors -->
<!-- ``` -->

<!-- We can manually set these categories as `levels` -->

<!-- ```{r} -->
<!-- pet_factors <- factor(pet_vec, levels = c("Dog", "Cat")) -->
<!-- ``` -->

<!-- ## Factors -->

<!-- - These are actually stored as integers -->
<!-- - Each integer corresponds to a `level` -->

<!-- ```{r, results='hide'} -->
<!-- str(pet_factors) -->
<!-- as.integer(pet_factors) -->
<!-- as.character(pet_factors) -->
<!-- ``` -->

<!-- ## Factors -->

<!-- **What would happen if we think a `factor` vector is a `character` vector, and we use it to select values from a `vector`/`matrix`/`data.frame`?** -->

<!-- ## Factors -->

<!-- **What would happen if we think a `factor` vector is a `character` vector, and we use it to select values from a `vector`/`matrix`/`data.frame`?** -->

<!-- ```{r} -->
<!-- names(pet_vec) <- pet_vec -->
<!-- pet_vec -->
<!-- pet_vec[pet_factors] -->
<!-- ``` -->

<!-- *This is why I'm very cautious about `read.csv()` and the standard `data.frame` etc.* -->

<!-- ## Factors -->

<!-- - There is a tidyverse package `forcats` -->
<!-- - A few improvements, particularly `fct_lump()` & `fct_infreq()` -->
<!--     - We can lump rare categories into `other` or order by frequency -->
